Causal Inference in the 2020 US Federal Election: While COVID-19 Lowered Trump's Support, Loss of Confidence Among Wealthier Voters Seems More Decisive

### 2023-12-08 outputs/paper/paper.qmd
Update the output paper

---
title: "Draft"
title-block-banner: true
author: "Yiliu Cao"
thanks: "Code and data from this analysis are available at: https://github.com/yiliuc/covid_and_trump_loss.git"
date: "today"
date-format: "long"
abstract: "TBD"
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

election_data <- read.csv(here("outputs/data/election_data_clean.csv"))
covid_data <- read.csv(here("outputs/data/covid_data_clean.csv"))
acs_data <- read.csv(here("outputs/data/acs_data_clean.csv"))
covid_election_data <- read.csv(here("outputs/data/covid_election.csv"))
data <- read.csv(here("outputs/data/merged_data.csv"))
election_2016_republican <- read.csv(here("outputs/data/election_data_2016.csv")) %>% 
  filter(party == "Republican") %>% 
  dplyr::select(fips, pct_vote_2016)

republican_data <- data %>% 
  filter(party == "Republican") %>% 
  inner_join(election_2016_republican, by = "fips") %>% 
  mutate(change_vote = pct_vote - pct_vote_2016)
```

@@

```{r}
#| echo: false
#| message: false
#| label: tbl-election
#| tbl-cap: The summary of voting patterns in 2020 US Federal Presidential Election
election_data %>% 
  group_by(party, candidate) %>% 
  summarise(total = sum(votes),
            mean = round(mean(pct_vote), 2),
            median = round(median(pct_vote), 2)) %>%
  arrange(desc(total)) %>% 
  rename(`Party` = party,
         `Candidate` = candidate,
         `Total Votes` = total,
         `Mean Pct Vote` = mean,
         `Median Pct Vote` = median) %>% 
  kable()
```

@@

```{r}
#| echo: false
#| message: false
#| label: tbl-covid
#| tbl-cap: The summary of COVID cases and deaths as of Election Day.
data %>% 
  group_by(winning_party) %>% 
  summarise(case = mean(cases),
            deaths = mean(deaths),
            inf = mean(infrate),
            mort = mean(mortrate),
            income = mean(mean_household_income)) %>% 
  rename(`Winning Party` = winning_party,
         `Cases` = case,
         `Deaths` = deaths,
         `Infection Rate` = inf,
         `Mortality Rate` = mort,
         `Income` = income) %>% 
  kable()
```

@@

data  %>%
  ggplot(mapping = aes(x = winning_party, y = mean_household_income)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.04, width = 0.3, height = 0) +
  theme_minimal() +
  labs(x = "Voting Party",
       y = "Death Rate per 1000")
```

@@

county <- map_data("county")
state <- map_data("state")
county_clean <- county %>% 
  mutate(state = region,
         county = subregion,
         state = get_state_abbreviation(toTitleCase(state)),
         county = toTitleCase(county)) %>% 
  filter(region != "AK") %>% 
  dplyr::select(-region, -subregion)

merged_data <- county_clean %>% left_join(republican_data, by = c("state", "county"))
```

@@

p1 <- ggplot(data = merged_data, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = ratio_rd), color = "black") +
  scale_fill_distiller(palette = "RdBu", direction = -1, na.value = "grey") +
  labs(fill = 'Share of Republican') +
  coord_quickmap() +
  theme_void() +
  geom_path(data = state, aes(x = long, y = lat, group = group), color = "black", linewidth = 0.5)

# Second plot
p2 <- ggplot(data = merged_data, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = infrate), color = "black") +
  scale_fill_gradient(low = "white", high = "darkblue", na.value = "grey",
                      limits = c(0, 5000), oob = scales::squish) +
  labs(fill = 'Infection Rate per 100k') +
  coord_quickmap() +
  theme_void() +
  geom_path(data = state, aes(x = long, y = lat, group = group), color = "black", linewidth = 0.5)

# Arrange the plots in a 2x1 layout
grid.arrange(p1, p2, ncol = 1)
```

@@

```{r}
#| echo: false
#| warning: false
#| message: false

highinf_model <- glm(treatment ~ prop_higher_education + pctile + 
                       no_insurance + private_insurance + males +
                       white_pct + black_pct, family = binomial(),
                      data = republican_data)
# summary(highinf_model)
republican_data$prop_score <- predict(highinf_model, type = "response")
```

@@

The above equation shows the logistic regression model to predict the propensity score for each county, given the covariates. It seems that insurance is a critical factor in the infection rate. Interestingly, the counties with a higher proportion of people with no insurance and private insurance will both increase the probability of this county having a high infection rate. Using this model, we can find the propensity score for each county and implement the three methods.

@@

library(Matching)
# Propensity score matching
rr <- Match(Y = republican_data$change_vote, Tr = republican_data$treatment, 
            X = republican_data$prop_score, M = 1)
# summary(rr)
```

```{r}
#| echo: false
#| message: false
#| label: tbl-PSS
#| tbl-cap: The results of causal effect using propensity score statification
# PSS
sub <- quantile(republican_data$prop_score, probs = c(0.2, 0.4, 0.6, 0.8))

sub1 <- glm(pct_vote ~ high_infrate, 
            data = republican_data[republican_data$prop_score <= sub[1], ])
sub2 <- glm(pct_vote ~ high_infrate, 
            data = republican_data[republican_data$prop_score > sub[1] & republican_data$prop_score <= sub[2], ])

sub3 <- glm(pct_vote ~ high_infrate, 
            data = republican_data[republican_data$prop_score > sub[2] & republican_data$prop_score <= sub[3], ])

sub4 <- glm(pct_vote ~ high_infrate, 
            data = republican_data[republican_data$prop_score > sub[3] & republican_data$prop_score <= sub[4], ])

sub5 <- glm(pct_vote ~ high_infrate, 
            data = republican_data[republican_data$prop_score > sub[4], ])

# Extracting the necessary information from each model
extract_info <- function(model, quantile) {
  coef_summary <- summary(model)$coefficients
  data.frame(
    quantile = quantile,
    ce = coef_summary["high_infrate", "Estimate"],
    p_val = coef_summary["high_infrate", "Pr(>|t|)"]
  )
}

@@

```{r}
#| echo: false
#| message: false
# Propensity score regression adjustment
model_adj <- glm(pct_vote ~ treatment + prop_score, data = republican_data)
# summary(model_adj)
```

```{r}
#| echo: false
#| message: false
# Counterfactual Analysis
adjusted_data <- data %>% 
  dplyr::select(state, county, fips, party, votes, total_votes, pct_vote, high_infrate)

@@

```{r}
#| echo: false
#| message: false
adjust_function2 <- function(data) {
  adjust_votes <- numeric(nrow(data))  # Initialize the vector with the correct length
  extra_vote_percentage <- 0.036928

>>>>

---
title: "Causal Inference in the 2020 US Federal Election: While COVID-19 Lowered Trump's Support, Loss of Confidence Among Wealthier Voters Seems More Decisive"
title-block-banner: true
author: "Yiliu Cao"
thanks: "Code and data from this analysis are available at: https://github.com/yiliuc/covid_and_trump_loss.git"
date: "today"
date-format: "long"
abstract: "This study used the election data from MIT EDSL and COVID-19 data from JHU CSSE, to investigate the causal inference between COVID-19 and Donald Trump's loss during the 2020 US Federal Election. The main methodology used in this paper is Propensity Score Matching with testing different treatments. The primary finding is that the counties with high death per case rate seems to vote less for Trump with approximate treatment effect -0.01. Besides, it seems that the interaction effect with income seems to be more critical: the medium-high income levels counties seems to vote at maximum 3% less for Trump compared to those low income counties. Furthermore, this paper also conduct counterfactual analysis and indicate the Trump would pre-elect if there was no COVID. Furture analysis should focus more on the choice of covariates to predict propensity scores."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

election_data <- read.csv(here("outputs/data/election_data_clean.csv"))
covid_data <- read.csv(here("outputs/data/covid_data_clean.csv"))
acs_data <- read.csv(here("outputs/data/acs_data_clean.csv"))
covid_election_data <- read.csv(here("outputs/data/covid_election.csv"))
data <- read.csv(here("outputs/data/merged_data.csv"))
```

@@

```{r}
#| echo: false
#| message: false
#| label: tbl-election
#| tbl-cap: The summary of voting patterns in 2020 US Federal Presidential Election
# election_data %>% 
#   group_by(party, candidate) %>% 
#   summarise(total = sum(votes),
#             mean = round(mean(pct_vote), 2),
#             median = round(median(pct_vote), 2)) %>%
#   arrange(desc(total)) %>% 
#   rename(`Party` = party,
#          `Candidate` = candidate,
#          `Total Votes` = total,
#          `Mean Pct Vote` = mean,
#          `Median Pct Vote` = median) %>% 
#   kable()
```

@@

```{r}
#| echo: false
#| message: false
#| label: tbl-covid
#| tbl-cap: The summary of COVID cases and deaths as of Election Day.
data %>% 
  group_by(winning_party) %>% 
  summarise(case = mean(cases),
            deaths = mean(deaths),
            inf = mean(infrate),
            dpc = mean(dpc),
            income = mean(mean_household_income)) %>% 
  rename(`Winning Party` = winning_party,
         `Cases` = case,
         `Deaths` = deaths,
         `Infection Rate` = inf,
         `Income` = income) %>% 
  kable()
```

@@

data  %>%
  ggplot(mapping = aes(x = winning_party, y = mean_household_income)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.04, width = 0.3, height = 0) +
  theme_minimal() +
  labs(x = "Winning Party",
       y = "Mean Household Income")
```

@@

county <- map_data("county")
state <- map_data("state")
county_clean <- county %>% 
  mutate(state = region,
         county = subregion,
         state = get_state_abbreviation(toTitleCase(state)),
         county = toTitleCase(county)) %>% 
  filter(region != "AK" & region != "HI") %>% 
  dplyr::select(-region, -subregion)

merged_data <- county_clean %>% 
  left_join(data, by = c("state", "county"))
```

```{r}
binned_data <- data %>%
  mutate(dpc_pctile = ntile(dpc, 100)) %>%  # Adjust 'breaks' to change number of bins
  group_by(dpc_pctile) %>%
  summarize(mean_rep = mean(pct_vote_rep, na.rm = TRUE),
            mean_demo = mean(pct_vote_demo, na.rm = TRUE)) %>% 
  mutate(bin = 1:n())

binned_data2 <- data %>%
  mutate(dpc_pctile = ntile(mean_household_income, 100)) %>%  # Adjust 'breaks' to change number of bins
  group_by(dpc_pctile) %>%
  summarize(mean_rep = mean(pct_vote_rep, na.rm = TRUE),
            mean_demo = mean(pct_vote_demo, na.rm = TRUE)) %>% 
  mutate(bin = 1:n(),
         sum = mean_rep + mean_demo)

# Plotting
ggplot(binned_data, aes(x = bin, y = mean_rep)) +
  geom_point() +  # or geom_line() depending on your preference
  geom_smooth(method = "lm",formula = y ~ poly(x, 4), se=FALSE) +
  theme_minimal() +
  labs(x = "Binned mortality per case ", y = "Mean vote for Trump",
       title = "Votes for Trump vs. MPC")
ggplot(binned_data, aes(x = bin, y = mean_demo)) +
  geom_point() +  # or geom_line() depending on your preference
  geom_smooth(method = "lm", formula = y ~ poly(x, 4),se=FALSE) +
  theme_minimal() +
  labs(x = "Binned mortality per case ", y = "Mean vote for Biden", 
       title = "Votes for Biden vs. MPC")

ggplot(binned_data2, aes(x = bin, y = mean_rep)) +
  geom_point() +  # or geom_line() depending on your preference
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se=FALSE) +
  theme_minimal() +
  labs(x = "Binned income levels", y = "Mean vote for Trump", 
       title = "Votes for Trump vs. Income")
ggplot(binned_data2, aes(x = bin, y = mean_demo)) +
  geom_point() +  # or geom_line() depending on your preference
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se=FALSE) +
  theme_minimal() +
  labs(x = "Binned income levels", y = "Mean vote for Biden",
       title = "Votes for Biden vs. Income")
```

@@

p1 <- ggplot(data = merged_data, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = pct_vote_rep), color = "black") +
  scale_fill_distiller(palette = "RdBu", direction = -1, na.value = "grey") +
  labs(title = "The relative share of votes between Republican and Democrat party",
       fill = 'Share of Republican') +
  coord_quickmap() +
  theme_void() +
  geom_path(data = state, aes(x = long, y = lat, group = group), color = "black", linewidth = 0.5)

# Second plot
p2 <- ggplot(data = merged_data, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = dpc), color = "black") +
  scale_fill_gradient(low = "white", high = "darkblue", na.value = "grey",
                      limits = c(0, 5000), oob = scales::squish) +
  labs(title = "The distribution of death per case rate by 100K",
       fill = 'Death per cases by 100k') +
  coord_quickmap() +
  theme_void() +
  geom_path(data = state, aes(x = long, y = lat, group = group), color = "black", linewidth = 0.5)

p3 <- ggplot(data = merged_data, aes(x = long, y = lat, group = group)) + 
  geom_polygon(aes(fill = mean_household_income), color = "black") +
  scale_fill_gradient(low = "white", high = "darkblue", na.value = "grey",
                      limits = c(60000, 120000), oob = scales::squish) +
  labs(title = "The distribution of mean household income",
       fill = 'Mean Hosehold Income') +
  coord_quickmap() +
  theme_void() +
  geom_path(data = state, aes(x = long, y = lat, group = group), color = "black", linewidth = 0.5)

# Arrange the plots in a 2x1 layout
grid.arrange(p1, p3, p2, ncol = 1)
```

```{r}
#| eval: false
library(sp)
library(geojsonio)
library(ggplot2)
library(dplyr)
library(broom)
hex_county <- geojson_read("us_county_hexgrid.geojson", what = "sp")

# Reformat the data as needed (depending on your data structure)
hex_county@data <- hex_county@data %>% 
  mutate(adjusted_field = gsub("some_pattern", "replacement", original_field))

# Fortify the data for ggplot
hex_county_fortified <- tidy(hex_county, region = "desired_region_field")

# Plot the hexagon map
ggplot() +
  geom_polygon(data = hex_county_fortified, aes(x = long, y = lat, group = group), fill="#a1dab4", color="#f7f7f7") +
  geom_text(data = hex_county_fortified, aes(x = long, y = lat, label = label_field), size = 3) +
  theme_void() +
  coord_map()
```

@@

```{r}
#| echo: false
#| warning: false
#| message: false
data <- read.csv(here("outputs/data/merged_data.csv"))

highinf_model <- glm(treatment ~ prop_higher_education + pctile + 
                       no_insurance + private_insurance + males +
                       white_pct + black_pct, family = binomial(),
                      data = data)
# summary(highinf_model)
data$prop_score <- predict(highinf_model, type = "response")
```

@@

The above equation shows the logistic regression model to predict the propensity score for each county, given the covariates. It seems that insurance is a critical factor in the infection rate. Interestingly, the counties with a higher proportion of people with no insurance and private insurance will both increase the probability of this county having a high infection rate. Using this model, we can find the propensity score for each county and implement the three methods.
```{r}
model <- lm(pct_vote_rep ~ dpc + mean_household_income, data = data)
summary(model)
```

@@

library(Matching)
# Propensity score matching
rr <- Match(Y = data$pct_vote_rep, Tr = data$treatment, 
            X = data$prop_score, M = 1)
summary(rr)
```

```{r}
#| echo: false
#| message: false
#| label: tbl-PSS
#| tbl-cap: The results of causal effect using propensity score statification
#| eval: false
# PSS
sub <- quantile(data$prop_score, probs = c(0.2, 0.4, 0.6, 0.8))

sub1 <- glm(pct_vote_rep ~ treatment,
            data = data[data$prop_score <= sub[1], ])
sub2 <- glm(pct_vote_rep ~ treatment,
            data = data[data$prop_score > sub[1] & data$prop_score <= sub[2], ])

sub3 <- glm(pct_vote_rep ~ treatment,
            data = data[data$prop_score > sub[2] & data$prop_score <= sub[3], ])

sub4 <- glm(pct_vote_rep ~ treatment,
            data = data[data$prop_score > sub[3] & data$prop_score <= sub[4], ])

sub5 <- glm(pct_vote_rep ~ treatment,
            data = data[data$prop_score > sub[4], ])

# Extracting the necessary information from each model
extract_info <- function(model, quantile) {
  coef_summary <- summary(model)$coefficients
  data <- data.frame(
    quantile = quantile,
    ce = coef_summary["treatment", "Estimate"],
    p_val = coef_summary["treatment", "Pr(>|t|)"]
  )
  return(data)
}

@@

```{r}
#| echo: false
#| message: false
# Propensity score regression adjustment
model_adj <- glm(change_vote_rep ~ treatment + prop_score, data = data)
summary(model_adj)
```

```{r}
#| echo: false
#| message: false
#| eval: false
# Counterfactual Analysis
adjusted_data <- data %>% 
  dplyr::select(state, county, fips, party, votes, total_votes, pct_vote, high_infrate)

@@

```{r}
#| echo: false
#| message: false
#| eval: false
adjust_function2 <- function(data) {
  adjust_votes <- numeric(nrow(data))  # Initialize the vector with the correct length
  extra_vote_percentage <- 0.036928
